Leadership across cultures and contexts: an exploratory analysis

### 2020-05-02 leadership_across_cultures_contexts.Rmd
Updates to cluster analysis results.

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic), with cultures labeled by subsistence strategy and represented by a symbol proportional to the number of documents that provided ethnographic texts for that culture (see also Figure \@ref(fig:subsisttree)). Our search yielded `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one variable in the study.

@@

Subsistence strategy was confounded with continental region. African text records were predominantly from horticultural societies, for instance, whereas North American text records were predominantly from hunter-gatherer societies (see Figure \@ref(fig:regionsubsistence)).

@@

We explored whether three features of our corpus of ethnographic texts -- date of publication, female gender of any author, and total pages of ethnography from each culture -- were associated with discovery of evidence for the `r length(all_study_vars)` variables in our study that coded for functions and qualities, benefits, and costs (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)). Dates of publication ranged from `r min(documents$d_publication_date)` to `r max(documents$d_publication_date)`, with a median year of `r median(documents$d_publication_date)` (see Figure \@ref(fig:doc-years)). There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine (some authors were co-authors). Of the `r nrow(documents)` source documents, `r sum(documents$female_coauthor)` (`r signif(100*sum(documents$female_coauthor)/nrow(documents), 2)`%) had a female author or co-author. Total pages of ethnography on each culture in the PSF ranged from `r min(leader_cult[['ehraf_pages']])` to `r as.integer(max(leader_cult[['ehraf_pages']]))`, with a median of `r median(leader_cult[['ehraf_pages']])`.

The number of text records on leadership retrieved for a particular culture was weakly correlated with the total number of pages of ethnography in the eHRAF (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). A loess regression suggested that the number of text records initially increased with the total number of pages of ethnography, plateauing at about 35 records around 4500 pages (see Figure \@ref(fig:eHRAFpages)).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary study variable as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but the publication date coefficient was significant in nine models. More recent publications were more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure \@ref(fig:pubdate). See supplementary information for details.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using logistic mixed effects models that controlled for one possible source of bias -- publication date -- as a fixed effect, and included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that variable (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Supplementary Table S??.

@@

The most common *functions* leaders serve, which are documented in over 70% of cultures, are resolving conflicts, providing various social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. Leaders also commonly serve as a representatives in inter-group interactions, manage the distribution of resources, and provide context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punish other group members and manage economic systems. Notable infrequent functions include controlling group movement and migrations.

The most common *qualities* of leaders are being of high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of the cultures sampled here leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained strong authority over followers. Notable infrequent qualities of leaders in these data include physical formidability and high social contacts (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

*High status* was the most frequent leader quality in the ethnographic record, with `r statusPercTxt`% of text records and `r statusPercCult`% cultures containing evidence that leaders were of high social status. To explore factors that distinguished these records from those with no evidence for status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with lasso penalty) of high status (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provide evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the terms with the strongest predictive values; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also strong predictors and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`* were also positive predictors. Notable terms which offered predictive value but with near zero coefficients included, *influence*, *authority*, *headman*, and *chief*. See Figure \@ref(fig:highstatusplot).

We then investigated if the proportion of text records with evidence for high status differed among the different levels of social and political organization, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastoralists societies were more likely to provide evidence for high status, as was South America and East Asia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

@@

Here we report the cluster analysis of variables that used the correlation distance metric ($1 - cor$), and the Ward agglomeration algorithm (for results from the other methods, see the SI). Figure \@ref(fig:cluster)A displays a dendrogram from the cluster analysis of the `r length(quality_vars)` leader quality variables and Figure \@ref(fig:cluster)B displays a dendrogram from the cluster analysis of the `r length(function_vars)` leader function variables. Each cluster analysis includes two estimates of significance for how strongly each cluster within the dendrogram is supported by the data. We rely on the AU (Approximately Unbiased) *p* values, which are computed by multiscale bootstrap resampling and represented as percentages (e.g., clusters with AU values $>95$ are strongly supported, and the top-level clusters are automatically highlighted by rectangles).

@@

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) identified two top-level clusters, one that comprised mostly variables related to material, social, and reproductive success, and one that comprised mostly variables indicating the relationship between leaders and their groups, which we term *social orientation*. Within the latter top-level cluster we identify two interpretable though weakly supported sub-clusters which are annotated with text. Within the first top-level cluster, we identify a subcluster mainly comprising variables related to the ability to provide social benefits, and a subcluster mainly comprising variables related to the ability to impose social costs.

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management functions cluster (i.e., *Organize cooperation*, *Military command*, *Resolve conflict*) and the smaller of which we identified as a prosociality cluster (i.e., *Social functions*, *Distribute resources*, *Hospitality*). Within the management cluster, we identified a mediation subcluster (e.g., *Resolve conflict*, *Provide counsel*), and an organization subcluster (e.g., *Organize cooperation*, *Military command*).

In naming the clusters and subclusters, we gave extra weight to variables that were strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes we identified.

We then fit separate logistic mixed effects models of each feature as a function of subsistence strategy and region, with random effects for author nested within culture. After adjusting the subsistence coefficient p-values for an FDR of 0.05, we computed and plotted the estimated marginal means of the significant models. See Figure \@ref(fig:features).

>>>>

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic), with cultures labeled by subsistence strategy and represented by a symbol proportional to the number of documents that provided ethnographic texts for that culture (see also Supplimentary Figure \@ref(fig:subsisttree)). Our search yielded `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one variable in the study.

@@

Subsistence strategy was confounded with continental region. African text records were predominantly from horticultural societies, for instance, whereas North American text records were predominantly from hunter-gatherer societies (see Supplimentary Figure \@ref(fig:regionsubsistence)).

@@

We explored whether three features of our corpus of ethnographic texts -- date of publication, female gender of any author, and total pages of ethnography from each culture -- were associated with discovery of evidence for the `r length(all_study_vars)` variables in our study that coded for functions and qualities, benefits, and costs (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)). Dates of publication ranged from `r min(documents$d_publication_date)` to `r max(documents$d_publication_date)`, with a median year of `r median(documents$d_publication_date)` (see Supplimentary Figure \@ref(fig:doc-years)). There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine (some authors were co-authors). Of the `r nrow(documents)` source documents, `r sum(documents$female_coauthor)` (`r signif(100*sum(documents$female_coauthor)/nrow(documents), 2)`%) had a female author or co-author. Total pages of ethnography on each culture in the PSF ranged from `r min(leader_cult[['ehraf_pages']])` to `r as.integer(max(leader_cult[['ehraf_pages']]))`, with a median of `r median(leader_cult[['ehraf_pages']])`.

The number of text records on leadership retrieved for a particular culture was weakly correlated with the total number of pages of ethnography in the eHRAF (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). A loess regression suggested that the number of text records initially increased with the total number of pages of ethnography, plateauing at about 35 records around 4500 pages (see Supplimentary Figure \@ref(fig:eHRAFpages)).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary study variable as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but the publication date coefficient was significant in nine models. More recent publications were more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Supplimentary Figure \@ref(fig:pubdate) and supplementary information for details.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support using logistic mixed effects models that controlled for one possible source of bias -- publication date -- as a fixed effect, and included authors nested within cultures as random intercepts. We then computed the proportion of cultures with at least one text record providing support for that variable (calculating 95% CI using a cluster bootstrap). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Supplementary Table \@ref(tab:glmmcoefs).

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing various social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. Leaders were also commonly described as serving as a representatives in inter-group interactions, managing the distribution of resources, and providing context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Notable infrequent functions included controlling group movement and migrations.

The most common qualities of leaders were having high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of the cultures sampled here leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained coercive authority over followers. Notable infrequent qualities of leaders in these data included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders were of high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplimentary information for details). We then fit an elasticnet logistic regression model (with lasso penalty) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the strongest predictive terms; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also relatively strong predictors and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were also positive predictors. Notable terms which offered predictive value but with near-zero coefficients included, *influence*, *authority*, *headman*, and *chief*. See Supplimentary Figure \@ref(fig:highstatusplot).

We then investigated if the proportion of text records with evidence for *High status* differed among the different levels of group structure, subsistence strategies, continental regions, and leader sex (again using a logistic mixed effects model with these variables as fixed effects and authors nested within cultures as random intercepts). There were modest differences. Hunter-gatherers and pastoralists were more likely to provide evidence for high status leaders, as was South America and East Asia. Female leaders were also more likely to be identified as high status. Somewhat counter-intuitively, the proportion of text records indicating high status leaders was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus). For the table of coefficients, see Supplementary Table X.

@@

Here we report the cluster analysis of variables that used the correlation distance metric ($1 - cor$), and the Ward agglomeration algorithm (for results from the other methods, see the supplimentary information). Figure \@ref(fig:cluster)A displays a dendrogram from the cluster analysis of the `r length(quality_vars)` leader quality variables and Figure \@ref(fig:cluster)B displays a dendrogram from the cluster analysis of the `r length(function_vars)` leader function variables. Each cluster analysis includes two estimates of significance for how strongly each cluster within the dendrogram is supported by the data. We rely on the AU (Approximately Unbiased) *p* values, which are computed by multiscale bootstrap resampling and represented as percentages (e.g., clusters with AU values $>95$ are strongly supported, and the top-level clusters are automatically highlighted by rectangles).

@@

The leader quality cluster analysis (Figure \@ref(fig:cluster)A) identified two top-level clusters, one that comprised mostly variables related to prosocial qualities and one that comprised other individual-level qualities which were not necessarily prosocial. Within the latter top-level cluster we identifed two interpretable and strongly supported subclusters which are annotated with text (*Social, reproductive, and material success* and *Competencies*). Within the first top-level cluster, we also identifed two subclusters, one of which contained meansures related to group-level pressures (*Cultural conformity*) and one which contined the prosocial traits (*Prosocial competencies*); these subclusters were moderately supported.

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management functions cluster (i.e., *Organize cooperation*, *Military command*, *Resolve conflict* functions) and the smaller of which we identified as a prosociality cluster (i.e., *Social functions*, *Distribute resources*, *Hospitality* functions). Within the management cluster, we identified two moderately suppported subclusters; a *Mediation* subcluster (e.g., *Resolve conflict*, *Provide counsel*), and an *Organization* subcluster (e.g., *Organize cooperation*, *Military command*).

We identify these named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature.

We then fit separate logistic mixed effects models of each feature as a function of subsistence strategy and region, with random effects for author nested within culture. After adjusting the subsistence coefficient p-values for an FDR of 0.05, we computed and plotted the estimated marginal means of the significant models. Two of these models (*Mediate* and *Prosocial competencies*) produced significant effects. See Figure \@ref(fig:features).
